---
title: "Final Project: PM 566"
subtitle: "University of Southern California, Fall 2024"
author: "Kendall Dimson"
---

## Description

Welcome to my PM 566 final project website, on the topic: "Analysing Rates and Trends of Vaccine Preventable Diseases in California, by County (2001-2022).

This project uses data from the California Department of Public Health (CDPH) and the Center for Disease Control and Prevention (CDC) to analyse trends of vaccine-preventable diseases in California. Vaccine-preventable diseases, as in the term name, are infectious diseases that can be prevented by vaccines. This project focused particularly on Pertussis (whooping cough), Invasive Meningococcal Disease, Hepatitis A, and COVID-19.

There are three main topics covered in this project:

  1. The project focuses on identifying previous vaccine-preventable disease epidemics in California. An epidemic is defined as "the occurrence of more cases of disease than expected in a given area or among a specific group of people over a particular period of time" (CDC).
  2. After identifying the vaccine-preventable disease with the highest total number of new cases over the years, the project investigated the distribution of the disease across California counties, adjusting for population density. Did the number of new cases vary in different regions? (This is to be explored in the full report)
  3. One of the main influential factors to the rise of new cases of vaccine-preventable disease is vaccine hesitancy. When regions do not have a high percentage of people vaccinated, then the community is susceptible to new infections of such vaccine-preventable cases. An example is with Measles, a highly infectious disease that can be prevented with the MMR vaccine. In recent years, regions throughout the U.S. have seen outbreaks of measle cases in regions with a lower vaccination rate. This project investigates if there is an association between vaccine hesitancy and new cases of vaccine-preventable diseases in California counties.

## Table 1: Total Cases of Vaccine-Preventable Diseases Reported in California, 2001-2022
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
vaccine <- read_csv("~/Desktop/a7a4d868-f6d2-49c9-8a11-904c69a0e3a0.csv")
county <- vaccine |> filter (County != 'California')
diseases_total <- county |> 
  group_by(Disease) |> 
  summarise(Total_Cases = sum(Count, na.rm = TRUE)) |> 
  arrange(desc(Total_Cases)) |> 
  slice_head(n = 15)

kable(diseases_total, 
      col.names = c("Disease", "Total Cases"))
```

## Figure 1: Bubble Map of Total Number of Pertussis cases in California (2001-2022)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Preliminary Analysis #2: Is there a geographical relationship between location and case incidence? 
#HeatMap of Pertussis
library (leaflet)
 pertussis <- county |> filter (Disease == 'Pertussis')
pertussis_tot <- pertussis |> group_by(County)|>
  summarise(Total_Cases =sum(Count, na.rm=TRUE)) |>
  ungroup()

county_coordinates <- data.frame(
  County = c(
    "Alameda", 
    "Alpine", 
    "Amador", 
    "Butte", 
    "Calaveras", 
    "Colusa", 
    "Contra Costa", 
    "Del Norte", 
    "El Dorado", 
    "Fresno", 
    "Glenn", 
    "Humboldt", 
    "Imperial", 
    "Inyo", 
    "Kern", 
    "Kings", 
    "Lake", 
    "Lassen", 
    "Los Angeles", 
    "Marin", 
    "Mariposa", 
    "Mendocino", 
    "Merced", 
    "Modoc", 
    "Mono", 
    "Monterey", 
    "Napa", 
    "Nevada", 
    "Orange", 
    "Placer", 
    "Plumas", 
    "Riverside", 
    "Sacramento", 
    "San Benito", 
    "San Bernardino", 
    "San Diego", 
    "San Francisco", 
    "San Joaquin", 
    "San Luis Obispo", 
    "San Mateo", 
    "Santa Barbara", 
    "Santa Clara", 
    "Santa Cruz", 
    "Shasta", 
    "Sierra", 
    "Siskiyou", 
    "Solano", 
    "Sonoma", 
    "Stanislaus", 
    "Sutter", 
    "Tehama", 
    "Trinity", 
    "Tulare", 
    "Tuolumne", 
    "Ventura", 
    "Yolo", 
    "Yuba"
  ),
  lat = c(
    37.6010,  # Alameda
    38.3652,  # Alpine
    38.3970,  # Amador
    39.4739,  # Butte
    38.2009,  # Calaveras
    39.2068,  # Colusa
    38.0016,  # Contra Costa
    41.7469,  # Del Norte
    38.7420,  # El Dorado
    36.7470,  # Fresno
    39.5051,  # Glenn
    40.6765,  # Humboldt
    32.6654,  # Imperial
    36.5784,  # Inyo
    35.4875,  # Kern
    36.0945,  # Kings
    38.9682,  # Lake
    40.5832,  # Lassen
    34.0522,  # Los Angeles
    38.0449,  # Marin
    37.4875,  # Mariposa
    39.4502,  # Mendocino
    37.3020,  # Merced
    41.5572,  # Modoc
    37.9500,  # Mono
    36.6777,  # Monterey
    38.2975,  # Napa
    39.2199,  # Nevada
    33.7175,  # Orange
    39.0910,  # Placer
    39.9724,  # Plumas
    33.9533,  # Riverside
    38.5816,  # Sacramento
    36.5772,  # San Benito
    34.1083,  # San Bernardino
    32.7157,  # San Diego
    37.7749,  # San Francisco
    37.9563,  # San Joaquin
    35.3733,  # San Luis Obispo
    37.5631,  # San Mateo
    34.4208,  # Santa Barbara
    37.3541,  # Santa Clara
    36.9741,  # Santa Cruz
    40.7906,  # Shasta
    39.5272,  # Sierra
    41.4936,  # Siskiyou
    38.2494,  # Solano
    38.5004,  # Sonoma
    37.4959,  # Stanislaus
    39.0194,  # Sutter
    40.0982,  # Tehama
    40.6542,  # Trinity
    36.1400,  # Tulare
    38.0291,  # Tuolumne
    34.3705,  # Ventura
    38.7647,  # Yolo
    39.1345   # Yuba
  ),
  lon = c(
    -122.2727,  # Alameda
    -119.8489,  # Alpine
    -120.6080,  # Amador
    -121.8076,  # Butte
    -120.5606,  # Calaveras
    -122.0015,  # Colusa
    -121.8058,  # Contra Costa
    -124.1018,  # Del Norte
    -120.5351,  # El Dorado
    -119.7724,  # Fresno
    -122.1740,  # Glenn
    -124.1958,  # Humboldt
    -115.3746,  # Imperial
    -116.2028,  # Inyo
    -118.8597,  # Kern
    -119.8817,  # Kings
    -122.8008,  # Lake
    -120.6548,  # Lassen
    -118.2437,  # Los Angeles
    -122.5156,  # Marin
    -119.9309,  # Mariposa
    -123.8060,  # Mendocino
    -120.4844,  # Merced
    -120.5420,  # Modoc
    -118.9560,  # Mono
    -121.6555,  # Monterey
    -122.2869,  # Napa
    -120.5734,  # Nevada
    -117.8310,  # Orange
    -120.8055,  # Placer
    -120.4841,  # Plumas
    -117.3962,  # Riverside
    -121.4944,  # Sacramento
    -121.5025,  # San Benito
    -116.5810,  # San Bernardino
    -117.1611,  # San Diego
    -122.4194,  # San Francisco
    -121.3134,  # San Joaquin
    -120.6544,  # San Luis Obispo
    -122.3710,  # San Mateo
    -119.6982,  # Santa Barbara
    -121.9552,  # Santa Clara
    -121.9552,  # Santa Cruz
    -122.4640,  # Shasta
    -119.5890,  # Sierra
    -122.4540,  # Siskiyou
    -122.0364,  # Solano
    -122.4064,  # Sonoma
    -120.8450,  # Stanislaus
    -121.6656,  # Sutter
    -122.4300,  # Tehama
    -123.1382,  # Trinity
    -118.8597,  # Tulare
    -120.2370,  # Tuolumne
    -119.1396,  # Ventura
    -121.9018,  # Yolo
    -121.5885   # Yuba
  )
)


heatmap_data <- pertussis_tot|>
  left_join(county_coordinates, by = "County")

leaflet(heatmap_data) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    weight = 1,
    radius = ~sqrt(Total_Cases) * 0.5,
    color = "green",
    fillOpacity = 0.5,
    popup = ~paste(County, ": ", Total_Cases, " cases")
  ) |>
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) 

```

## Figure 2: Trends of Pertussis Incidence in Los Angeles (2001-2022)


```{r, echo=FALSE, message=FALSE, warning=FALSE}
 pertussis <- county |> filter (Disease == 'Pertussis')

 #Stratify Pertussis, Check Los Angeles Trend
pertussis_la <- pertussis |> 
  filter(County == 'Los Angeles')
library(dplyr)

# Aggregate total cases by year
annual_cases <- pertussis_la %>%
  group_by(Year) %>%
  summarise(Total_Cases = sum(Count, na.rm = TRUE))

library(ggplot2)

# Line graph for Total Cases per Year
ggplot(annual_cases, aes(x = Year, y = Total_Cases)) +
  geom_line(color = "darkgreen", linewidth = 1.5) +
  geom_point(color = "darkgreen", linewidth = 3) +
  labs(title = "Figure 2. Trend of Pertussis Cases in Los Angeles per Year",
       x = "Year",
       y = "Total Cases") +
  theme_minimal()

```
In Figure 2, we investigated the trend of Pertussis cases in Los Angeles each year, the county who had the most cumulative incidence for Pertussis. It was found that Los Angeles county had peaks of case incidence in 2010, 2014, and 2019. After 2020 there was a lower decline of Pertussis cases, but this is likely due to the occurrence of the COVID-19 pandemic.

```{r,echo=FALSE, message=FALSE, warning=FALSE}
  #Stratify IMD, Check LA Trend
IMD <- county |> filter (Disease == 'Invasive Meningococcal Disease')
IMD_kern <- IMD |> 
  filter(County == 'Kern')

ggplot(IMD_kern, aes(x = Year, y = Count)) +
  geom_bar(stat = "identity", fill = "plum") +
  labs(title = "Figure 3: Invasive Menginococcal Cases in Kern County, by Year",
       x = "Year",
       y = "Number of Cases") +
  theme_minimal() 
```
In Figure 3, one of the interesting findings was that there was an Invasive Meningococcal Disease epidemic that occurred in Kern County, a rural region in Central California. The spread of new cases does not follow the trend of new cases that occurred in Los Angeles County and San Diego county (futher explored in full report).